---
title: "KOA-Mindset Secondary Analyses"
output: html_document
date: "2026-08-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set-up

First, all necessary libraries are imported, the data is read in, and all necessary variables are produced.

```{r set_up, message=FALSE}
# First, the libraries 
# If you want to replicate my environment exactly, use renv::restore()
library(tidyverse)
library(mediation)
library(lavaan)
library(multcomp)
library(knitr)
library(lme4)
select = dplyr::select

# Next, reading the data
GLAB = c("Mindset", "Education", "No-intervention")
df = read.csv("../deidentified_data.csv") %>%
  mutate(group = recode_values(t2_Group.ID, from = c(1, 2, 3),
                               to = GLAB), # Group names
         female = case_when(t1_sex == 1 ~ 1,
                            t1_sex == 0 ~ 0,
                            TRUE ~ NA), # Recode sex
         chg_PEM_avg = t2_PEM_avg - t1_PEM_avg, # Change scores
         t3_OAcomp = t3_IMI_man + t3_IMI_opp - t3_IMI_cat,
         t2_OAcomp = t2_IMI_man + t2_IMI_opp - t2_IMI_cat,
         t1_OAcomp = t1_IMI_man + t1_IMI_opp - t1_IMI_cat,
         Y_surgery = t3_surgery - t1_surgery,
         Y_limit = case_when(t1_management_11 - t3_management_11 > 0 ~ 1,
                             TRUE ~ 0), # Only improvement is a "1"
         Y_inject = case_when(t1_management_2 - t3_management_2 > 0 ~ 1,
                              TRUE ~ 0),
         Y_TSK = t3_TSK_avg - t1_TSK_avg,
         Y_ASEp = t3_ASE_pain_avg - t1_ASE_pain_avg,
         chg_OAcomp = t2_OAcomp - t1_OAcomp, # Baseline features
         bilateral = case_when(t1_uni.bi == 5 ~ 1,
                               t1_uni.bi == 4 ~ 0,
                               TRUE ~ NA),
         minority = case_when(!(t1_race %in% c("6", "")) ~ 1,
                              TRUE ~ 0),
         employed = case_when(grepl("[1-2]", t1_employment) ~ 1,
                              TRUE ~ 0),
         depression = case_when(grepl("10", t1_med.cond) ~ 1, 
                                TRUE ~ 0),
         n_comorbid = lengths(strsplit(t1_med.cond, ",")),
         Y_pase = t3_PASE_total - t1_PASE_total) %>%
  mutate(group = factor(group, levels = GLAB[c(3,2,1)]))
```

# Analysis 1 - Moderators of the Treatment Effect

## Calculate p-values

For the moderation analysis, we map over all variables of interest, estimate the corresponding model, and extract the interaction/p-values from the result. 

```{r moderation_models}
# Variables of interest
vars = c("female", "minority", "employed", "bilateral", "t1_age",
         "t1_bmi", "n_comorbid", "t1_pain", "t1_pain.time", "t1_ASE_pain_avg",
         "t1_TSK_avg", "t1_PASE_total")
names(vars) = c("Female", "Racial/ethnic minority", "Employed", "Bilateral symptoms",
          "Age", "BMI", "# comorbidities",  "Baseline Pain", "Symptom duration",
          "Baseline self-efficacy", "Baseline fear of movement",
          "Baseline activity")

# Standardize continuous variables, convert binary to factor
bin_vars = vars[1:4]
cont_vars = vars[-c(1:4)]
for_mod = df %>%
  mutate(across(all_of(cont_vars), ~ scale(.x)[,1]),
         across(all_of(bin_vars), ~ factor(.x)))

analysis_type = "primary"
mod_string = switch(analysis_type, 
                    primary = "t2_PEM_avg ~ t1_PEM_avg + group*", 
                    adjust_1 = "t2_PEM_avg ~ t1_PEM_avg + t1_pain + group*", 
                    adjust_2 = "t2_PEM_avg ~ t1_PEM_avg + t1_pain + depression + group*", 
                    adjust_3 = "t2_PEM_avg ~ t1_PEM_avg + t1_pain + depression + t1_ASE_function_avg + t1_ASE_pain_avg + group*") 

# Map over variables and fit models, conditional on baseline mindset
model_fits = map(vars, function(x){
  # Model formula
  form = as.formula(paste0(mod_string, x)) 
  
  # Fit OLS regression
  mod = lm(form, data = for_mod) 
  
  # Extract coefficients and p-values
  coef_df = summary(mod)$coefficients 
  
  # Estimand is in the last row
  return(data.frame(p_value = coef_df[nrow(coef_df), 4],
                    interaction = coef_df[nrow(coef_df), 1]))
}) %>% list_rbind()
```

## Adjusting p-values

We next adjust the p-values to account for multiple testing. Several different adjustment routines are considered as sensitivity analysis, though the main findings use Benjamini Hochberg. Bonferroni correction, Benjamini Yukeiteli, and maxT all are known to be more conservative than Benjamini Hochberg. As the maxT procedure is not available from a library in R, it is implemented from scratch below. A good resource for this procedure is [here](https://brainder.org/2017/01/18/why-the-maximum-statistic/).

```{r moderation_maxT}
n_boot = 10000

raw_t_stats = map(vars, function(x){
  # Model formula
  form = as.formula(paste0(mod_string, x))
  
  # Fit the model
  mod = lm(form, data = for_mod)
  
  # Extract the coefficients
  coef_df = summary(mod)$coefficients
  
  # Estimand is in last row, t-statistic is in third column
  return(coef_df[nrow(coef_df), 3])
}) %>% unlist()

# Next, permute the predictors (keeping the responses fixed), and estimate
# max t-statistics
permute_maxT = map(1:n_boot, function(x){
  # Permute the dataset
  permuted = for_mod[sample(nrow(for_mod), replace = F),]
  permuted$t2_PEM_avg = for_mod$t2_PEM_avg

  # Calculate t-statistics for permuted dataset
  perm_t = map(vars, function(y){
    # Model formula
    form = as.formula(paste0(mod_string, y))
    
    # Fit the model
    mod = lm(form, data = permuted)
    
    # Extract the coefficients
    coef_df = summary(mod)$coefficients
    
    # Estimand is in last row, t-statistic is in third column
    return(coef_df[nrow(coef_df), 3])
  }) %>% unlist()

  # Return the maximum absolute t-statistic among the predictors
  return(max(abs(perm_t)))
}) %>% unlist()

# maxT-adjusted p-value is the empirical quantile of the absolute t-statistics
# within the distribution of maximum absolute t-statistics
maxT_corrected = map(1:length(vars), function(idx){
  return(mean(abs(raw_t_stats[idx]) <= permute_maxT))
}) %>% unlist()
```

We now compare these p-values with the other methods.

```{r moderation_pvals}
data.frame(Var = names(vars),
           interaction = model_fits$interaction,
           p_value = model_fits$p_value,
           BH = p.adjust(model_fits$p_value, method = "BH"),
           BY = p.adjust(model_fits$p_value, method = "BY"),
           Bonferroni = p.adjust(model_fits$p_value, method = "bonferroni"),
           MaxT = maxT_corrected) %>%
  arrange(p_value)
```

Baseline activity and self-efficacy show up as consistent moderators, regardless of adjustment method.  Baseline fear of movement and baseline pain are also significant using Benjamini-Hochberg adjustment, but they lose moderator status under more conservative adjustment routines.

## Evaluating the Benjamini-Hochberg Procedure

To quickly assess whether the covariate structure in this case conflicts with the assumptions of the Benjamini-Hochberg procedure (tests are independent or exhibit positive regression dependence), we perform a simple empirical assessment. By permuting the subject's covariate values while keeping the responses fixed, we break all associations between predictors and outcomes. We can then estimate the moderation models, apply the Benjamini-Hochberg procedure, and assess the false discovery rate for the tests now known to be null. As all tests are truly null, note that Benjamini-Hochberg is expected to be much more conservative, nearly on-par with Bonferroni in control of the Family-wise Error Rate (rather than the False Discovery Rate).

```{r moderation_eval}
# Number of bootstrap replicates
n_boot = 10000

false_discovery = map(1:n_boot, function(x){
  # Permute the dataset
  permuted = for_mod[sample(nrow(for_mod), replace = F),]
  permuted$t2_PEM_avg = for_mod$t2_PEM_avg
  
  # P-values after permutation
  perm_pvals = map(vars, function(y){
    # Model formula
    form = as.formula(paste0("t2_PEM_avg ~ t1_PEM_avg + group*", y))
    
    # Fit the model on the permuted data
    mod = lm(form, data = permuted)
    
    # Extract model coefficients
    coef_df = summary(mod)$coefficients
    
    # Estimand is in last row, p-value is in fourth column
    return(coef_df[nrow(coef_df), 4])
  }) %>% unlist()
  
  # Apply the Benjamini-Hochberg adjustment
  adj_pvals = p.adjust(perm_pvals, method = "BH")

  # Calculate FDR
  return(mean(adj_pvals < 0.05))
}) %>% unlist()

mean(false_discovery) 
```

The average false discovery rate is considerably below the desired 0.05 level. 

# Analysis 2 - Mindset Mediation of Treatment -> Activity

## Mediation Analysis

We perform a causal mediation analysis to assess change in exercise mindset (time point 2 - time point 1) as a potential mediator of the treatment effect on activity change (time point 3 - time point 1). . Rather that including the changes as predictors/outcomes directly, we include the latter measures adjusted for the earlier ones. Uncertainty is assessed using bootstrap inference.

```{r mind_med_activity}
# Filter data to mindset vs no intervention
fit_df = df %>%
  filter(t2_Group.ID %in% c(1, 3)) %>%
  mutate(Treat = case_when(t2_Group.ID == 1 ~ 1,
                           TRUE ~ 0))

# Mediator model
model1 = lm(t2_PEM_avg ~ t1_PEM_avg + Treat, data = fit_df)

# Direct + indirect effect model
model2 = lm(t3_PASE_total ~ t1_PASE_total + t2_PEM_avg + t1_PEM_avg + Treat, data = fit_df)

# Mediation analysis
med_res = mediate(model1, model2, treat = "Treat", mediator = "t2_PEM_avg",
                  boot = TRUE, sims = 1000)
summary(med_res)
```

The indirect effect (ACME) is just 6.81, not significantly different from zero at a 0.05 level. The proportion of the activity effect explained by the mindset treatment is estimated to be 26\%. 

## Correlation Assessment

To perform a rough assessment of why this might be the case, we can examine the correlation between mindset change (time point 2 - time point 1) and activity change (time point 3 - time point 1) to see whether there is evidence of a relationship between the mediator and the outcome.

```{r mind_med_check}
cor(fit_df$chg_PEM_avg, fit_df$Y_pase)
```

The low correlation is indicative of a weak relationship between these two measures, likely interrupting the indirect mediation path through mindset change.

# Analysis 3 - Mindset Mediation of Treatment -> Mindset Outcomes

## Mediation

We perform separate causal mediation analyses to assess mindset change (time point 2 - time point 1) as a potential mediator of the treatment effect on mindset-specific outcomes. These outcomes include difference in perceived need for surgery (time point 3 - time point 1), removal of self-imposed physical limitation (time point 1 to time point 3), and removal of injection use as a management strategy (time point 1 to time point 3). Generalized linear models are used to model the last two (binary) outcomes. Uncertainty is assessed using bootstrap inference.

We begin by assessing mediation of the MINDSET vs NONE treatment effect.

```{r mind_med_outcomes_vs_no, message = FALSE}
# Outcomes of interest
outcomes = c("t3_surgery", "Y_limit", "Y_inject")
out_names = c("Perceived need for surgery",
              "Imposing physical limitations",
              "Injection Use")

# Filter data to mindset vs no intervention
fit_df_vs_none = df %>%
  filter(t2_Group.ID %in% c(1, 3)) %>%
  mutate(Treat = case_when(t2_Group.ID == 1 ~ 1,
                           TRUE ~ 0))

output_vs_none = data.frame()
for(idx in 1:length(outcomes)){
  # Variable name
  varname = outcomes[idx]
  
  # Mediator model
  model1 = glm(t2_OAcomp ~ t1_OAcomp + Treat, data = fit_df_vs_none, family = "gaussian")
  
  # Direct and indirect effects model (binary outcome)
  form_string = "~ t2_OAcomp + t1_OAcomp + Treat"
  if(varname == "t3_surgery"){ form_string = paste0(form_string, " + t1_surgery") }
  form = as.formula(paste0(varname, form_string))
  fam = ifelse(varname %in% c("Y_limit", "Y_inject"), "binomial", "gaussian")
  model2 = glm(form, data = fit_df_vs_none, family = fam)
  
  # Mediation analysis
  med_res = mediate(
    model1,
    model2,
    treat = "Treat",
    mediator = "t2_OAcomp",
    boot = TRUE,
    sims = 1000
  )
  summ_med = summary(med_res)
  
  output_vs_none = rbind(output_vs_none,
                 data.frame(outcome = out_names[idx],
                      indirect = summ_med$d0,
                      CI_LB = unname(summ_med$d0.ci[1]),
                      CI_UB = unname(summ_med$d0.ci[2]),
                      p_value = summ_med$d0.p,
                      Percent_mediated = summ_med$n0*100))
}
output_vs_none
```

We see that mindset mediates the treatment effect (MINDSET vs NONE) on both "perceived need for surgery" and "imposing physical limitations", but not on "injection use". Note that the "percent_mediated" column just reports the relative size of the mediation effect relative to the total effect, so it can fall outside of the 0\%-100\% range. 

We can continue by similarly assessing mediation of the MINDSET vs EDUCATION treatment effect for the same outcomes.

```{r mind_med_outcomes_vs_education, message = FALSE}
# Filter data to mindset vs education
fit_df_vs_edu = df %>%
  filter(t2_Group.ID %in% c(1, 2)) %>%
  mutate(Treat = case_when(t2_Group.ID == 1 ~ 1,
                           TRUE ~ 0))

output_vs_edu = data.frame()
for(idx in 1:length(outcomes)){
  # Variable name
  varname = outcomes[idx]
  
  # Mediator model
  model1 = glm(t2_OAcomp ~ t1_OAcomp + Treat, data = fit_df_vs_edu, family = "gaussian")
  
  # Direct and indirect effects model (binary outcome)
  form_string = "~ t2_OAcomp + t1_OAcomp + Treat"
  if(varname == "t3_surgery"){ form_string = paste0(form_string, " + t1_surgery") }
  form = as.formula(paste0(varname, form_string))
  fam = ifelse(varname %in% c("Y_limit", "Y_inject"), "binomial", "gaussian")
  model2 = glm(form, data = fit_df_vs_edu, family = fam)
  
  # Mediation analysis
  med_res = mediate(
    model1,
    model2,
    treat = "Treat",
    mediator = "t2_OAcomp",
    boot = TRUE,
    sims = 1000
  )
  summ_med = summary(med_res)
  
  output_vs_edu = rbind(output_vs_edu,
                 data.frame(outcome = out_names[idx],
                      indirect = summ_med$d0,
                      CI_LB = unname(summ_med$d0.ci[1]),
                      CI_UB = unname(summ_med$d0.ci[2]),
                      p_value = summ_med$d0.p,
                      Percent_mediated = summ_med$n0*100))
}
output_vs_edu
```

We see that mindset meaningfully mediates the treatment effect (MINDSET vs EDUCATION) on perceived need for surgery here as well, in a similar capacity as was observed for the effect of MINDSET vs NONE.

## Multiple Testing Adjustment

We can perform a simple multiple testing adjustment of the 6 raw p-values so far estimated, grouped together as part of examining the potential exercise mindset mediation of the treatment effect(s) on mindset-specific outcomes. 

```{r mind_med_padj}
# Join outputs of the previous two analyses
view_df = rbind(output_vs_none %>% 
                  select(outcome, p_value) %>%
                  mutate(Reference = "None"), 
                output_vs_edu %>% 
                  select(outcome, p_value) %>%
                  mutate(Reference = "Education"))

# Adjust the p-values using Benjamini-Hochberg and Bonferroni
view_df$BH = p.adjust(view_df$p_value, method = "BH")
view_df$Bonferroni = p.adjust(view_df$p_value, method = "bonferroni")

# Visualize
view_df %>%
  arrange(p_value)
```

After Benjamini-Hochberg correction, there is still significant mediation of the MINDSET vs NONE treatment effect for the outcomes of "imposing physical limitations" and "perceived need for surgery", as well as of the MINDSET vs EDUCATION treatment effect for the "perceived need for surgery" outcome at the 0.05 level. Under the more conservative Bonferroni adjustment, the mediation of the MINDSET vs EDUCATION treatment effect on the "perceived need for surgery" outcome is no longer significant at this level.

# Analysis 4 - Mindset Mediation of Treatment -> Psychosocial Outcomes

## Mediation

We perform separate causal mediation analyses to assess mindset change (time point 2 - time point 1) as a potential mediator of the treatment effect on psycho-social outcomes. These outcomes include fear of movement change and change in pain self-efficacy (time point 3 - time point 1). Uncertainty is estimated using bootstrap inference.

We begin by assessing mediation of the MINDSET vs NONE treatment effect.

```{r mind_med_psych_vs_no, message = FALSE}
# Outcomes of interest
outcomes = c("Y_TSK", "Y_ASEp")
out_names = c("Fear of movement",
              "Pain self-efficacy")

# Filter data to mindset vs no intervention
fit_df_vs_none = df %>%
  filter(t2_Group.ID %in% c(1, 3)) %>%
  mutate(Treat = case_when(t2_Group.ID == 1 ~ 1,
                           TRUE ~ 0))

output_vs_none = data.frame()
for(idx in 1:length(outcomes)){
  # Variable name
  varname = outcomes[idx]
  
  # Mediator model
  model1 = lm(chg_OAcomp ~ Treat, data = fit_df_vs_none)
  
  # Direct and indirect effects model (binary outcome)
  form = as.formula(paste0(varname, "~ chg_OAcomp + Treat"))
  model2 = lm(form, data = fit_df_vs_none)
  
  # Mediation analysis
  med_res = mediate(
    model1,
    model2,
    treat = "Treat",
    mediator = "chg_OAcomp",
    boot = TRUE,
    sims = 1000
  )
  summ_med = summary(med_res)
  
  output_vs_none = rbind(output_vs_none,
                 data.frame(outcome = out_names[idx],
                      indirect = summ_med$d0,
                      CI_LB = unname(summ_med$d0.ci[1]),
                      CI_UB = unname(summ_med$d0.ci[2]),
                      p_value = summ_med$d0.p,
                      Percent_mediated = summ_med$n0*100))
}
output_vs_none
```

We see that mindset meaningfully mediates the treatment effect (MINDSET vs NONE) on both "fear of movement" and "pain self-efficacy".

We can continue by similarly assessing mediation of the MINDSET vs EDUCATION treatment effect for the same outcomes.

```{r mind_med_psych_vs_edu, message = FALSE}
# Filter data to mindset vs education
fit_df_vs_edu = df %>%
  filter(t2_Group.ID %in% c(1, 2)) %>%
  mutate(Treat = case_when(t2_Group.ID == 1 ~ 1,
                           TRUE ~ 0))

output_vs_edu = data.frame()
for(idx in 1:length(outcomes)){
  # Variable name
  varname = outcomes[idx]
  
  # Mediator model
  model1 = lm(chg_OAcomp ~ Treat, data = fit_df_vs_edu)
  
  # Direct and indirect effects model (binary outcome)
  form = as.formula(paste0(varname, "~ chg_OAcomp + Treat"))
  model2 = lm(form, data = fit_df_vs_edu)
  
  # Mediation analysis
  med_res = mediate(
    model1,
    model2,
    treat = "Treat",
    mediator = "chg_OAcomp",
    boot = TRUE,
    sims = 1000
  )
  summ_med = summary(med_res)
  
  output_vs_edu = rbind(output_vs_edu,
                 data.frame(outcome = out_names[idx],
                      indirect = summ_med$d0,
                      CI_LB = unname(summ_med$d0.ci[1]),
                      CI_UB = unname(summ_med$d0.ci[2]),
                      p_value = summ_med$d0.p,
                      Percent_mediated = summ_med$n0*100))
}
output_vs_edu
```

The mediation is similarly strong when considering the MINDSET vs EDUCATION treatment effect, with similar proportions of mediation.

## Multiple Testing Adjustment

We can again perform a multiple-testing adjustment, grouping together the 4 tests (2 outcomes x 2 treatment effects) as part of examining the potential exercise mindset mediation of the treatment effect(s) on psychosocial outcomes. 

```{r mind_med_padj}
# Join outputs of the previous two analyses
view_df = rbind(output_vs_none %>% 
                  select(outcome, p_value) %>%
                  mutate(Reference = "None"), 
                output_vs_edu %>% 
                  select(outcome, p_value) %>%
                  mutate(Reference = "Education"))

# Adjust the p-values using Benjamini-Hochberg and Bonferroni
view_df$BH = p.adjust(view_df$p_value, method = "BH")
view_df$Bonferroni = p.adjust(view_df$p_value, method = "bonferroni")

# Visualize
view_df %>%
  arrange(p_value)
```

After Benjamini-Hochberg correction, all mediations are still significant at the 0.05 level. Even under the much more conservative Bonferroni correction, all mediations are significant at the same level other than the mediation of the MINDSET vs EDUCATION treatment effect on the "fear of movement" outcome.

# Analysis 5 - Serial Mediation of Mindset -> Self-efficacy -> Activity

We will examine a two-step mediation model, where treatment affects mindset, which affects self-efficacy, which finally affects the activity outcome. All predictors earlier in this chain are assumed to be able to directly influence the downstream variables/outcomes. We focus on examining the relative magnitudes of the indirect and direct effects of mindset on the activity outcome. This model can be fit using structural equation modeling software as follows, where inference is accomplished using bootstrap inference. First, we focus on the treatment effect of MINDSET vs NONE.

```{r serial_SEM_none}
# Filter data to mindset vs no intervention
for_mod_no = df %>%
  filter(t2_Group.ID %in% c(1, 3)) %>%
  mutate(Treat = case_when(t2_Group.ID == 1 ~ 1,
                           TRUE ~ 0), 
         Y_pase = scale(Y_pase)[,1]) # Scale for numerical stability

# Specify the structural equations model
serial_model = "
  chg_OAcomp ~ a * Treat
  Y_ASEp ~ b * chg_OAcomp + c * Treat
  Y_pase ~ d * Y_ASEp + e * chg_OAcomp + f * Treat
  ind_eff := d * b
"

# Fit the model
fit = sem(model = serial_model, data = for_mod_no, se = "boot",
          bootstrap = 5000)
parameterEstimates(fit, boot.ci.type = "norm") %>%
  filter(label %in% c("e", "ind_eff")) %>%
  select(label, est, pvalue, ci.lower, ci.upper) %>%
  mutate(label = case_when(label == "e" ~ "Direct Effect", 
                           TRUE ~ "Indirect Effect"))
```

From the above table, it is clear that the estimated direct effect is nearly zero, while the indirect effect is relatively much more positive. However, neither the direct nor indirect effect of mindset on activity is significant at the 0.05 level when considering treatment to be the MINDSET vs NONE contrast. 

We can repeat this analysis, instead focusing on the treatment effect of MINDSET vs EDUCATION.

```{r serial_SEM_edu}
# Filter data to mindset vs education
for_mod_edu = df %>%
  filter(t2_Group.ID %in% c(1, 2)) %>%
  mutate(Treat = case_when(t2_Group.ID == 1 ~ 1,
                           TRUE ~ 0),
         Y_pase = scale(Y_pase)[,1]) # Scale for numerical stability

fit = sem(model = serial_model, data = for_mod_edu, se = "boot",
          bootstrap = 5000)
parameterEstimates(fit, boot.ci.type = "norm") %>%
  filter(label %in% c("e", "ind_eff")) %>%
  select(label, est, pvalue, ci.lower, ci.upper) %>%
  mutate(label = case_when(label == "e" ~ "Direct Effect", 
                           TRUE ~ "Indirect Effect"))
```

When the treatment is adjusted to MINDSET vs EDUCATION, the direct effect becomes negative, though still not statistically significant from zero at a 0.05 level. The indirect effect, which is still positive under this treatment contrast, is now just barely statistically significant at the 0.05 level. Notice that, regardless of the treatment specification, the indirect effect is quite small on the scale of standard deviation of the activity outcome.

# Analysis 6 - Heterogeneity of Treatment Effects

We begin by defining a simple function of baseline knee pain (+), physical activity (-), and arthritis pain self-efficacy (-) which can be used as a suitability criterion. We begin by examining estimated treatment effect within each strata of this criterion.

```{r strata_simple_tertiles}
# Filter to MINDSET vs NONE
suit_df = df %>%
  filter(t2_Group.ID %in% c(1, 3)) %>% 
  mutate(Treat = case_when(t2_Group.ID == 1 ~ "treat",
                           TRUE ~ "no"))

# Define the criterion
suit_df = suit_df %>%
  mutate(z_t1_pain = scale(t1_pain)[,1], # Z-score contributing features
         z_t1_PASE_total = scale(t1_PASE_total)[,1],
         z_t1_ASE_pain_avg = scale(t1_ASE_pain_avg)[,1],
         suitability = z_t1_pain - z_t1_PASE_total - z_t1_ASE_pain_avg)

# Stratify and examine ATE (mean difference due to RCT)
suit_df %>%
  mutate(tertile = ntile(suitability, 3), # Generate tertiles
         Strata = case_when(tertile == 1 ~ "Low",
                            tertile == 2 ~ "Mid",
                            tertile == 3 ~ "High"), 
         Strata = factor(Strata, levels = c("Low", "Mid", "High"))) %>%
  group_by(Strata, Treat) %>%  # Calculate average outcome within groups
  summarise(avg_SE = mean(Y_ASEp), .groups = "drop_last") %>%
  pivot_wider(names_from = Treat, values_from = avg_SE) %>%
  mutate(ATE = treat - no) %>%
  select(-c(treat, no))
```

We can also examine the estimated treatment effects on self-efficacy within sub-populations chosen based upon thresholds for each of the contributing baseline features, giving an idea of the potential effects of enrollment rules. 

```{r strata_simple_enrollment}
ATE_est <- function(df){
  output = df %>%
    group_by(Treat) %>%
    summarise(avg_SE = mean(Y_ASEp)) %>%
    pivot_wider(names_from = Treat, values_from = avg_SE) %>%
    mutate(ATE = treat - no) %>%
    pull(ATE)
  return(output)
}

enroll_df = suit_df %>%
  mutate(pain_thresh = case_when(t1_pain >= median(t1_pain) ~ TRUE, TRUE ~ FALSE),
         selfeff_thresh = case_when(t1_ASE_pain_avg <= median(t1_ASE_pain_avg) ~ TRUE, 
                                    TRUE ~ FALSE), 
         activ_thresh = case_when(t1_PASE_total <= median(t1_PASE_total) ~ TRUE, 
                                  TRUE ~ FALSE))

highpain_lowselfeff = enroll_df %>%
  filter(pain_thresh & selfeff_thresh)
allthree = enroll_df %>%
  filter(pain_thresh & selfeff_thresh & activ_thresh)

data.frame(Enrollment_Rules = c("All", "High pain & low self-efficacy", "All three"), 
           Benefit = c(ATE_est(enroll_df),
                       ATE_est(highpain_lowselfeff),
                       ATE_est(allthree)),  
           Qualifying_Percent = 100 * c(1,
                                        nrow(highpain_lowselfeff) / nrow(enroll_df),
                                        nrow(allthree) / nrow(enroll_df)))
```

As an alternative to this fixed weighting, we can calculate the first principal component among these three baseline measures, which defines the axis of greatest variation among them for this population. Similar to the procedure used for the simple criterion, we can  examine the estimated treatment effect within strata of this new, PCA-based criterion.

```{r strata_pca}
# Calculate the first principal component
suit_df$PC1 = princomp(suit_df[,c("z_t1_pain", 
                                  "z_t1_PASE_total",
                                  "z_t1_ASE_pain_avg")])$scores[,1]

# Stratify and examine ATE (mean difference due to RCT)
suit_df %>%
  mutate(tertile = ntile(PC1, 3), # Generate tertiles
         Strata = case_when(tertile == 1 ~ "Low PC1",
                            tertile == 2 ~ "Mid PC1",
                            tertile == 3 ~ "High PC1"), 
         Strata = factor(Strata, levels = c("Low PC1", "Mid PC1", "High PC1"))) %>%
  group_by(Strata, Treat) %>%  # Calculate average outcome within groups
  summarise(avg_SE = mean(Y_ASEp), .groups = "drop_last") %>%
  pivot_wider(names_from = Treat, values_from = avg_SE) %>%
  mutate(ATE = treat - no) %>%
  select(-c(treat, no))
```

We actually see very similar differences in the stratified ATE estimates according to the first principal component, indicating that the simpler suitability metric is likely capturing much of the same information.

# Analysis 7 - Durability of Treatment Effects

We can model the impact of the treatment on exercise mindset and the adaptive OA composite longitudinally over time. This allows us to assess the impact of the treatment at time point 2 (post) and whether this effect persists at time point 3 (1-month after). 

```{r longitud_data}
# Orient the data appropriately
fit_df = inner_join(
  # Extract treatment group and the exercise mindset at all 3 time points per subject
  df %>%
    select(contains("PEM_avg"), group) %>%
    mutate(ID = row_number()) %>%
    select(-c("chg_PEM_avg")) %>%
    pivot_longer(-c(ID, group), names_to = "time", values_to = "Exercise Mindset") %>%
    mutate(time = substring(time, 2, 2), time = as.numeric(time)),
  # Extract treatment group and the OA composite at all 3 time points per subject
  df %>%
    select(contains("OAcomp"), group) %>%
    mutate(ID = row_number()) %>%
    select(-c("chg_OAcomp")) %>%
    pivot_longer(-c(ID, group), names_to = "time", values_to = "Adaptive OA Composite") %>%
    mutate(time = substring(time, 2, 2), time = as.numeric(time), 
           tp = case_when(time == 1 ~ "Baseline", 
                          time == 2 ~ "Post",
                          TRUE ~ "1-month"), # Convert time to categories
           tp = factor(tp, levels = c("Baseline", "Post", "1-month"))) 
)
```

We first visualize the longitudinal effects, first aggregating by treatment arm.

```{r longitud_vis_agg, message=FALSE}
# Calculate arm-level summaries at each time point
agg_df = fit_df %>%
  pivot_longer(-c(ID, time, tp, group), names_to = "Measure", values_to = "Val") %>%
  group_by(group, tp, Measure) %>%
  summarise(avg = mean(Val), se = sd(Val)/sqrt(n()))

# Create the visualization
agg_df %>%
  ggplot(aes(x = tp, color = group)) +
  geom_line(aes(y = avg, group = group), position = position_dodge(width = 0.2)) + 
  geom_pointrange(aes(y = avg, 
                      ymin = avg - 1.96*se, 
                      ymax = avg + 1.96*se), position = position_dodge(width = 0.2)) + 
  theme_classic() +
  facet_wrap(.~Measure, scales = "free_y") +
  labs(x = "Time Point", y = "", color = "")
```

We can now show these effects within the context of the individual level data to give an idea of the substantial heterogeneity in trajectories.

```{r longitud_vis_ind}
fit_df %>%
  pivot_longer(-c(ID, time, tp, group), names_to = "Measure", values_to = "Val") %>%
  ggplot(aes(x = tp, color = group)) +
  geom_line(aes(y = Val, group = ID), alpha = 0.1, position = position_jitter(width = 0.1)) +
  geom_line(data = agg_df, 
            aes(y = avg, group = group), 
            position = position_dodge(width = 0.2)) + 
  geom_pointrange(data = agg_df, 
                  aes(y = avg, 
                      ymin = avg - 1.96*se, 
                      ymax = avg + 1.96*se), position = position_dodge(width = 0.2)) + 
  theme_classic() +
  facet_wrap(.~Measure, scales = "free_y") +
  labs(x = "Time Point", y = "", color = "")
```

To handle variability in subjects beginning positions, we will fit two mixed effects models, one for the adaptive OA composite and one for exercise mindset. Predictors for this model will include an interaction between group and time point, and each model will incorporate subject-specific random effects. 

```{r longitud_model}
mod_exercise = lmer(`Exercise Mindset` ~ tp*group + (1|ID), data = fit_df)
mod_OAcomp = lmer(`Adaptive OA Composite` ~ tp*group + (1|ID), data = fit_df)
```

We can use these model fits, along with the corresponding linear contrasts, to estimate the gain from baseline to post-intervention, the subsequent degredation moving to 1-month post, and the overall effect moving from baseline to 1-month post intervention.

```{r longitud_lincomp}
# Function for calculating linear contrasts with CI and p-values (unadjusted)
calc_contrast = function(model, K){
  contrast = glht(model, linfct = K)
  out_df = data.frame(confint(contrast)$confint)
  out_df$p_value = summary(contrast)$test$pvalues
  out_df$Arm = c("No intervention", "Education", "Mindset")
  return(out_df)
}

# Estimate time point 1 to time point 2 contrasts
K_t1t2 = rbind(t(c(0, 1, 0, 0, 0, 0, 0, 0, 0)), # No intervention arm
                t(c(0, 1, 0, 0, 0, 1, 0, 0, 0)), # Education arm
                t(c(0, 1, 0, 0, 0, 0, 0, 1, 0))) # Mindset arm
ex_t1t2 = calc_contrast(mod_exercise, K_t1t2) %>% 
  mutate(Outcome = "Exercise Mindset", 
         Contrast = "T2 - T1")
oa_t1t2 = calc_contrast(mod_OAcomp, K_t1t2) %>%
  mutate(Outcome = "Adaptive OA Composite", 
         Contrast = "T2 - T1")

# Estimate time point 2 to time point 3 contrasts
K_t2t3 = rbind(t(c(0, -1, 1, 0, 0, 0, 0, 0, 0)), # No intervention arm
               t(c(0, -1, 1, 0, 0, -1, 1, 0, 0)), # Education arm
               t(c(0, -1, 1, 0, 0, 0, 0, -1, 1))) # Mindset arm
ex_t2t3 = calc_contrast(mod_exercise, K_t2t3) %>% 
  mutate(Outcome = "Exercise Mindset", 
         Contrast = "T3 - T2")
oa_t2t3 = calc_contrast(mod_OAcomp, K_t2t3) %>% 
  mutate(Outcome = "Adaptive OA Composite", 
         Contrast = "T3 - T2")

# Estimate time point 1 to time point 3 contrasts
K_t1t3 = rbind(t(c(0, 0, 1, 0, 0, 0, 0, 0, 0)), # No intervention arm
               t(c(0, 0, 1, 0, 0, 0, 1, 0, 0)), # Education arm
               t(c(0, 0, 1, 0, 0, 0, 0, 0, 1))) # Mindset arm
ex_t1t3 = calc_contrast(mod_exercise, K_t1t3) %>% 
  mutate(Outcome = "Exercise Mindset", 
         Contrast = "T3 - T1")
oa_t1t3 = calc_contrast(mod_OAcomp, K_t1t3) %>% 
  mutate(Outcome = "Adaptive OA Composite", 
         Contrast = "T3 - T1")

# Combine for output
output_df = rbind(ex_t1t2, oa_t1t2, ex_t2t3, oa_t2t3, ex_t1t3, oa_t1t3) %>%
  mutate(Contrast = factor(Contrast, levels = c("T2 - T1", "T3 - T2", "T3 - T1")))

# Exercise Mindset
output_df %>%
  filter(Outcome == "Exercise Mindset")%>%
  arrange(Outcome, Arm, Contrast)

# Adaptive OA composite
output_df %>%
  filter(Outcome == "Adaptive OA Composite")%>%
  arrange(Outcome, Arm, Contrast)
```

We see that regressions are common - the T3 - T2 contrasts are negative and significant at the 0.05 level for the Mindset arm (p-values and CI multiple-testing adjusted, with grouping by outcome and time point contrast). However, the overall contrasts (T3 - T1) remain positive and significant at the same level.

We can now explore whether baseline pain and fear of movement impact the decaying relationship by adding them (after binarization at the median) as modifiers to a model of similar structure but restricted to those within the MINDSET treatment arm.

```{r longitud_modifiers}
# Add in modifiers
fit_df = inner_join(fit_df, df %>% mutate(ID = row_number()) %>% select(ID, t1_pain, t1_TSK_avg)) %>%
  mutate(base_pain = case_when(t1_pain >= median(t1_pain) ~ 1, 
                                                 TRUE ~ 0), 
         base_pain = factor(base_pain), 
         base_fear_movt = case_when(t1_TSK_avg >= median(t1_TSK_avg) ~ 1, 
                                    TRUE ~ 0), 
         base_fear_movt = factor(base_fear_movt))

# Fit the modification models
exercise_pain = lmer(`Exercise Mindset` ~ tp * base_pain  + (1 | ID), data = fit_df %>% filter(group == "Mindset"))
OAcomp_pain = lmer(`Adaptive OA Composite` ~ tp * base_pain + (1|ID), data = fit_df %>% filter(group == "Mindset"))
exercise_fear = lmer(`Exercise Mindset` ~ tp * base_fear_movt  + (1 | ID), data = fit_df %>% filter(group == "Mindset"))
OAcomp_fear = lmer(`Adaptive OA Composite` ~ tp * base_fear_movt + (1|ID), data = fit_df %>% filter(group == "Mindset"))

# Contrasts
K = matrix(c(0, 0, 0, 0, -1, 1), nrow = 1)
```

We begin with the effect of baseline pain on exercise mentality.

```{r cont1}
summary(glht(exercise_pain, linfct = K))
```


Next, for the effect of baseline pain on the adaptive OA composite.

```{r cont2}
summary(glht(OAcomp_pain, linfct = K))
```

Next, we examine the effect of baseline fear of movement on exercise mentality.

```{r cont3}
summary(glht(exercise_fear, linfct = K))
```

Finally, we estimate the effect of baseline fear on the adaptive OA composite.

```{r cont4}
summary(glht(OAcomp_fear, linfct = K))
```

We see that the contrasts of interest, indicating the difference in the decay for those above the median values of baseline pain and baseline fear of movement. We find substantial negative difference (faster decay) for both groups when the outcome is exercise mentality, but no substantial results for the Adaptive OA composite models.
