---
title: "Treatment Pathways"
author: "Joe Sartini"
date: "2026-09-01"
output:
  html_document: default
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set-up

First, all necessary libraries are imported, the data is read in, and all necessary variables are produced.

```{r set_up, message=FALSE}
# First, the libraries 
# If you want to replicate my environment exactly, use renv::restore()
library(tidyverse)
library(mediation)
library(lavaan)
library(multcomp)
library(knitr)
library(lme4)
select = dplyr::select

# Next, reading the data
GLAB = c("Mindset", "Education", "No-intervention")
df = read.csv("../deidentified_data.csv") %>%
  mutate(group = recode_values(t2_Group.ID, from = c(1, 2, 3),
                               to = GLAB), # Group names
         group = factor(group, levels = GLAB[c(3,2,1)]),
         female = case_when(t1_sex == 1 ~ 1,
                            t1_sex == 0 ~ 0,
                            TRUE ~ NA), # Recode sex
         t3_OAcomp = t3_IMI_man + t3_IMI_opp - t3_IMI_cat,
         t2_OAcomp = t2_IMI_man + t2_IMI_opp - t2_IMI_cat,
         t1_OAcomp = t1_IMI_man + t1_IMI_opp - t1_IMI_cat,
         Y_surgery = t3_surgery - t1_surgery,
         Y_limit = case_when(t1_management_11 - t3_management_11 > 0 ~ 1, # Improvement is a "1"
                             t1_management_11 == 0 ~ NA, # Don't consider those not using the strategy
                             TRUE ~ 0),  # Participants who used the strategy and did not improve
         Y_TSK = t3_TSK_avg - t1_TSK_avg,
         Y_ASEp = t3_ASE_pain_avg - t1_ASE_pain_avg,
         bilateral = case_when(t1_uni.bi == 5 ~ 1,
                               t1_uni.bi == 4 ~ 0,
                               TRUE ~ NA),
         minority = case_when(!(t1_race %in% c("6", "")) ~ 1,
                              TRUE ~ 0),
         employed = case_when(grepl("[1-2]", t1_employment) ~ 1,
                              TRUE ~ 0),
         depression = case_when(grepl("10", t1_med.cond) ~ 1, 
                                TRUE ~ 0),
         n_comorbid = lengths(strsplit(t1_med.cond, ",")))
```

# Mindset Mediation of Treatment -> Mindset Outcomes

## Mediation

We perform separate causal mediation analyses to assess mindset change (time point 2 - time point 1) as a potential mediator of the treatment effect on mindset-specific outcomes. These outcomes include difference in perceived need for surgery (time point 3 - time point 1) and removal of self-imposed physical limitation (time point 1 to time point 3). Generalized linear models are used to model the last two (binary) outcomes. Uncertainty is assessed using bootstrap inference.

We begin by assessing mediation of the MINDSET vs NONE treatment effect.

```{r mind_med_outcomes_vs_no, message = FALSE}
# Outcomes of interest
outcomes = c("t3_surgery", "Y_limit")
out_names = c("Perceived need for surgery",
              "Imposing physical limitations")

# Filter data to mindset vs no intervention
fit_df_vs_none = df %>%
  filter(t2_Group.ID %in% c(1, 3)) %>%
  mutate(Treat = case_when(t2_Group.ID == 1 ~ 1,
                           TRUE ~ 0))

output_vs_none = data.frame()
for(idx in 1:length(outcomes)){
  # Variable name
  varname = outcomes[idx]
  
  # Mediator model
  if(varname == "Y_limit"){
    fit_df = fit_df_vs_none %>%
      filter(!is.na(Y_limit))
  }
  else{
    fit_df = fit_df_vs_none
  }
  model1 = glm(t2_OAcomp ~ t1_OAcomp + Treat, data = fit_df, family = "gaussian")
  
  # Direct and indirect effects model (binary outcome)
  form_string = "~ t2_OAcomp + t1_OAcomp + Treat"
  if(varname == "t3_surgery"){ form_string = paste0(form_string, " + t1_surgery") }
  form = as.formula(paste0(varname, form_string))
  fam = ifelse(varname == "Y_limit", "binomial", "gaussian")
  model2 = glm(form, data = fit_df, family = fam)
  
  # Mediation analysis
  med_res = mediate(
    model1,
    model2,
    treat = "Treat",
    mediator = "t2_OAcomp",
    boot = TRUE,
    sims = 1000
  )
  summ_med = summary(med_res)
  
  output_vs_none = rbind(output_vs_none,
                 data.frame(outcome = out_names[idx],
                      indirect = summ_med$d0,
                      CI_LB = unname(summ_med$d0.ci[1]),
                      CI_UB = unname(summ_med$d0.ci[2]),
                      p_value = summ_med$d0.p,
                      Percent_mediated = summ_med$n0*100))
}
output_vs_none
```

We see that mindset mediates the treatment effect (MINDSET vs NONE) on both "perceived need for surgery" and "imposing physical limitations". Note that the "percent_mediated" column just reports the relative size of the mediation effect relative to the total effect, so it can fall outside of the 0\%-100\% range. 

We can continue by similarly assessing mediation of the MINDSET vs EDUCATION treatment effect for the same outcomes.

```{r mind_med_outcomes_vs_education, message = FALSE}
# Filter data to mindset vs education
fit_df_vs_edu = df %>%
  filter(t2_Group.ID %in% c(1, 2)) %>%
  mutate(Treat = case_when(t2_Group.ID == 1 ~ 1,
                           TRUE ~ 0))

output_vs_edu = data.frame()
for(idx in 1:length(outcomes)){
  # Variable name
  varname = outcomes[idx]
  
  if(varname == "Y_limit"){
    fit_df = fit_df_vs_edu %>%
      filter(!is.na(Y_limit))
  }
  else{
    fit_df = fit_df_vs_edu
  }
  
  # Mediator model
  model1 = glm(t2_OAcomp ~ t1_OAcomp + Treat, data = fit_df, family = "gaussian")
  
  # Direct and indirect effects model (binary outcome)
  form_string = "~ t2_OAcomp + t1_OAcomp + Treat"
  if(varname == "t3_surgery"){ form_string = paste0(form_string, " + t1_surgery") }
  form = as.formula(paste0(varname, form_string))
  fam = ifelse(varname == "Y_limit", "binomial", "gaussian")
  model2 = glm(form, data = fit_df, family = fam)
  
  # Mediation analysis
  med_res = mediate(
    model1,
    model2,
    treat = "Treat",
    mediator = "t2_OAcomp",
    boot = TRUE,
    sims = 1000
  )
  summ_med = summary(med_res)
  
  output_vs_edu = rbind(output_vs_edu,
                 data.frame(outcome = out_names[idx],
                      indirect = summ_med$d0,
                      CI_LB = unname(summ_med$d0.ci[1]),
                      CI_UB = unname(summ_med$d0.ci[2]),
                      p_value = summ_med$d0.p,
                      Percent_mediated = summ_med$n0*100))
}
output_vs_edu
```

We see that mindset mediates the treatment effects (MINDSET vs EDUCATION) on both outcomes here as well, though the "imposing physical limitations" outcome is mediated much less.

## Multiple Testing Adjustment

We can perform a simple multiple testing adjustment of the 6 raw p-values so far estimated, grouped together as part of examining the potential exercise mindset mediation of the treatment effect(s) on mindset-specific outcomes. 

```{r mind_med_padj}
# Join outputs of the previous two analyses
view_df = rbind(output_vs_none %>% 
                  select(outcome, p_value) %>%
                  mutate(Reference = "None"), 
                output_vs_edu %>% 
                  select(outcome, p_value) %>%
                  mutate(Reference = "Education"))

# Adjust the p-values using Benjamini-Hochberg and Bonferroni
view_df$BH = p.adjust(view_df$p_value, method = "BH")
view_df$Bonferroni = p.adjust(view_df$p_value, method = "bonferroni")

# Visualize
view_df %>%
  arrange(p_value)
```

After Benjamini-Hochberg and Bonferroni correction, there is still significant mediation of the MINDSET vs NONE treatment effect for both outcomes, as well as of the MINDSET vs EDUCATION treatment effect for the "perceived need for surgery" outcome at the 0.05 level. 

# Mindset Mediation of Treatment -> Psychosocial Outcomes

## Mediation

We perform separate causal mediation analyses to assess mindset change (time point 2 - time point 1) as a potential mediator of the treatment effect on psycho-social outcomes. These outcomes include fear of movement change and change in pain self-efficacy (time point 3 - time point 1). Uncertainty is estimated using bootstrap inference.

We begin by assessing mediation of the MINDSET vs NONE treatment effect.

```{r mind_med_psych_vs_no, message = FALSE}
# Outcomes of interest
outcomes = c("t3_TSK_avg", "t3_ASE_pain_avg", "t3_ASE_function_avg")
out_names = c("Fear of movement",
              "Pain self-efficacy", 
              "Other symptom self-efficacy")

# Filter data to mindset vs no intervention
fit_df_vs_none = df %>%
  filter(t2_Group.ID %in% c(1, 3)) %>%
  mutate(Treat = case_when(t2_Group.ID == 1 ~ 1,
                           TRUE ~ 0))

output_vs_none = data.frame()
for(idx in 1:length(outcomes)){
  # Variable name
  varname = outcomes[idx]
  
  # Mediator model
  model1 = lm(t2_OAcomp ~ t1_OAcomp + Treat, data = fit_df_vs_none)
  
  # Direct and indirect effects model (binary outcome)
  form_string = "~ t2_OAcomp + t1_OAcomp + Treat"
  if(varname == "t3_TSK_avg"){
    form_string = paste0(form_string, " + t1_TSK_avg")
  }
  else if(varname == "t3_ASE_pain_avg"){
    form_string = paste0(form_string, " + t1_ASE_pain_avg")
  }
  else{
    form_string = paste0(form_string, " + t1_ASE_function_avg")
  }
  form = as.formula(paste0(varname, form_string))
  model2 = lm(form, data = fit_df_vs_none)
  
  # Mediation analysis
  med_res = mediate(
    model1,
    model2,
    treat = "Treat",
    mediator = "t2_OAcomp",
    boot = TRUE,
    sims = 1000
  )
  summ_med = summary(med_res)
  
  output_vs_none = rbind(output_vs_none,
                 data.frame(outcome = out_names[idx],
                      indirect = summ_med$d0,
                      CI_LB = unname(summ_med$d0.ci[1]),
                      CI_UB = unname(summ_med$d0.ci[2]),
                      p_value = summ_med$d0.p,
                      Percent_mediated = summ_med$n0*100))
}
output_vs_none
```

We see that mindset meaningfully mediates the treatment effect (MINDSET vs NONE) on both "fear of movement" and "pain self-efficacy".

We can continue by similarly assessing mediation of the MINDSET vs EDUCATION treatment effect for the same outcomes.

```{r mind_med_psych_vs_edu, message = FALSE}
# Filter data to mindset vs education
fit_df_vs_edu = df %>%
  filter(t2_Group.ID %in% c(1, 2)) %>%
  mutate(Treat = case_when(t2_Group.ID == 1 ~ 1,
                           TRUE ~ 0))

output_vs_edu = data.frame()
for(idx in 1:length(outcomes)){
  # Variable name
  varname = outcomes[idx]
  
  # Mediator model
  model1 = lm(t2_OAcomp ~ t1_OAcomp + Treat, data = fit_df_vs_edu)
  
  # Direct and indirect effects model (binary outcome)
  form_string = "~ t2_OAcomp + t1_OAcomp + Treat"
  if(varname == "t3_TSK_avg"){
    form_string = paste0(form_string, " + t1_TSK_avg")
  }
  else if(varname == "t3_ASE_pain_avg"){
    form_string = paste0(form_string, " + t1_ASE_pain_avg")
  }
  else{
    form_string = paste0(form_string, " + t1_ASE_function_avg")
  }
  form = as.formula(paste0(varname, form_string))
  model2 = lm(form, data = fit_df_vs_edu)
  
  # Mediation analysis
  med_res = mediate(
    model1,
    model2,
    treat = "Treat",
    mediator = "t2_OAcomp",
    boot = TRUE,
    sims = 1000
  )
  summ_med = summary(med_res)
  
  output_vs_edu = rbind(output_vs_edu,
                 data.frame(outcome = out_names[idx],
                      indirect = summ_med$d0,
                      CI_LB = unname(summ_med$d0.ci[1]),
                      CI_UB = unname(summ_med$d0.ci[2]),
                      p_value = summ_med$d0.p,
                      Percent_mediated = summ_med$n0*100))
}
output_vs_edu
```

The mediation is similarly strong when considering the MINDSET vs EDUCATION treatment effect, with similar, slightly higher, proportions of mediation.

## Multiple Testing Adjustment

We can again perform a multiple-testing adjustment, grouping together the 4 tests (2 outcomes x 2 treatment effects) as part of examining the potential exercise mindset mediation of the treatment effect(s) on psychosocial outcomes. 

```{r psych_med_padj}
# Join outputs of the previous two analyses
view_df = rbind(output_vs_none %>% 
                  select(outcome, p_value) %>%
                  mutate(Reference = "None"), 
                output_vs_edu %>% 
                  select(outcome, p_value) %>%
                  mutate(Reference = "Education"))

# Adjust the p-values using Benjamini-Hochberg and Bonferroni
view_df$BH = p.adjust(view_df$p_value, method = "BH")
view_df$Bonferroni = p.adjust(view_df$p_value, method = "bonferroni")

# Visualize
view_df %>%
  arrange(p_value)
```

After Benjamini-Hochberg correction, all mediation p-values are incredibly small, indicating significance at the 0.05 level. Even under the much more conservative Bonferroni correction, all mediations are significant at the same level.
