---
title: "Treatment Pathways"
author: "Joe Sartini"
date: "2026-09-01"
output:
  html_document: default
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set-up

First, all necessary libraries are imported, the data is read in, and all necessary variables are produced.

```{r set_up, message=FALSE}
# First, the libraries 
# If you want to replicate my environment exactly, use renv::restore()
library(tidyverse)
library(mediation)
library(lavaan)
library(knitr)
library(ggtext)
select = dplyr::select

# Next, reading the data
GLAB = c("Mindset", "Education", "No-intervention")
df = read.csv("../deidentified_data.csv") %>%
  mutate(group = recode_values(t2_Group.ID, from = c(1, 2, 3),
                               to = GLAB), # Group names
         group = factor(group, levels = GLAB[c(3,2,1)]),
         female = case_when(t1_sex == 1 ~ 1,
                            t1_sex == 0 ~ 0,
                            TRUE ~ NA), # Recode sex
         t3_OAcomp = t3_IMI_man + t3_IMI_opp - t3_IMI_cat,
         t2_OAcomp = t2_IMI_man + t2_IMI_opp - t2_IMI_cat,
         t1_OAcomp = t1_IMI_man + t1_IMI_opp - t1_IMI_cat,
         Y_surgery = t3_surgery - t1_surgery,
         Y_limit = case_when(t1_management_11 - t3_management_11 > 0 ~ 1, # Improvement is a "1"
                             t1_management_11 == 0 ~ NA, # Don't consider those not using the strategy
                             TRUE ~ 0),  # Participants who used the strategy and did not improve
         Y_TSK = t3_TSK_avg - t1_TSK_avg,
         Y_ASEp = t3_ASE_pain_avg - t1_ASE_pain_avg,
         bilateral = case_when(t1_uni.bi == 5 ~ 1,
                               t1_uni.bi == 4 ~ 0,
                               TRUE ~ NA),
         minority = case_when(!(t1_race %in% c("6", "")) ~ 1,
                              TRUE ~ 0),
         employed = case_when(grepl("[1-2]", t1_employment) ~ 1,
                              TRUE ~ 0),
         depression = case_when(grepl("10", t1_med.cond) ~ 1, 
                                TRUE ~ 0),
         n_comorbid = lengths(strsplit(t1_med.cond, ",")))
```

# Primary Analyses: OA Mindset Mediation of 5 Outcomes

## Mediation

We perform separate causal mediation analyses to assess mindset change (time point 2 - time point 1) as a potential mediator of the treatment effect on 5 outcomes which were impacted by the treatment. These outcomes include difference in perceived need for surgery (time point 3 - time point 1), removal of self-imposed physical limitation (time point 1 to time point 3), fear of movement change, change in pain self-efficacy (time point 3 - time point 1), and change in other symptom self-efficacy (time point 3 - time point 1). 

We assess mediation of the movement from both the no intervention and education arms to the mindset treatment arm. Generalized linear models are used to model the binary removal of self-imposed physical limitation outcome. Uncertainty is assessed using bootstrap inference.

```{r mediate_primary, message = FALSE}
# Outcomes of interest
outcomes = c("t3_surgery", "Y_limit", "t3_TSK_avg", "t3_ASE_pain_avg", "t3_ASE_function_avg")
out_names = c("Perceived need for surgery",
              "Imposing physical limitations", 
              "Fear of movement",
              "Pain self-efficacy", 
              "Other symptom self-efficacy")

primary_df = data.frame()
for(idx in 1:length(outcomes)){
  varname = outcomes[idx]
  
  # Data create
  if(varname == "Y_limit"){
    fit_df = df %>%
      filter(!is.na(Y_limit))
  }
  else{
    fit_df = df
  }
  
  # Mediator model
  model1 = glm(t2_OAcomp ~ t1_OAcomp + group, data = fit_df, family = "gaussian")
  
  # Direct and indirect effects model (binary outcome)
  add_form = switch(varname, 
                    t3_surgery = "+ t1_surgery",
                    t3_TSK_avg = "+ t1_TSK_avg",
                    t3_ASE_pain_avg = "+ t1_ASE_pain_avg",
                    t3_ASE_function_avg = "+ t1_ASE_function_avg", 
                    Y_limit = "")
  form_string = paste0("~ t2_OAcomp + t1_OAcomp + group", add_form)
  
  form = as.formula(paste0(varname, form_string))
  fam = ifelse(varname == "Y_limit", "binomial", "gaussian")
  
  model2 = glm(form, data = fit_df, family = fam)
  
  # Mediation analysis
  contrast_no = mediate(
    model1,
    model2,
    treat = "group",
    mediator = "t2_OAcomp",
    control.value = "No-intervention",
    treat.value = "Mindset",
    boot = T,
    sims = 100000,
    parallel = "multicore",
    ncpus = 8
  )
  
  summ_no = summary(contrast_no)
  output_no = data.frame(
    total = summ_no$tau.coef,
    direct = summ_no$z0,
    indirect = summ_no$d0,
    CI_LB = unname(summ_no$d0.ci[1]),
    CI_UB = unname(summ_no$d0.ci[2]),
    p_value = summ_no$d0.p,
    Percent_mediated = summ_no$n0 * 100, 
    N_obs = summ_no$nobs
  )
  
  contrast_edu = mediate(
    model1,
    model2,
    treat = "group",
    mediator = "t2_OAcomp",
    control.value = "Education",
    treat.value = "Mindset",
    boot = T,
    sims = 100000, 
    parallel = "multicore",
    ncpus = 8
  )
  
  summ_edu = summary(contrast_edu)
  output_edu = data.frame(
    total = summ_edu$tau.coef,
    direct = summ_edu$z0,
    indirect = summ_edu$d0,
    CI_LB = unname(summ_edu$d0.ci[1]),
    CI_UB = unname(summ_edu$d0.ci[2]),
    p_value = summ_edu$d0.p,
    Percent_mediated = summ_edu$n0 * 100, 
    N_obs = summ_edu$nobs
  )
  
  outcome_df = rbind(
    output_no %>% mutate(Comparison = "No intervention"),
    output_edu %>% mutate(Comparison = "Education")
  ) %>%
    mutate(Outcome = out_names[idx])
  
  primary_df = rbind(primary_df, outcome_df)
}

primary_df %>%
  select(Comparison, Outcome, total, direct, indirect, CI_LB, CI_UB, p_value, Percent_mediated, N_obs) %>%
  arrange(Comparison, Outcome)
```

We see that mindset mediates the treatment effects for both contrasts and all outcomes quite strongly. Note that the "percent_mediated" column just reports the relative size of the mediation effect relative to the total effect, so it can fall outside of the 0\%-100\% range. 

## Multiple Testing Adjustment

We can perform a simple multiple testing adjustment of the 10 raw p-values so far estimated, grouped together as part of examining the potential exercise mindset mediation of the treatment effect(s) the chosen outcomes. 

```{r mult_adjust_primary}
# Join outputs of the previous two analyses
view_df = primary_df %>%
  select(Comparison, Outcome, p_value)

# Adjust the p-values using Benjamini-Hochberg and Bonferroni
view_df$BH = p.adjust(view_df$p_value, method = "BH")
view_df$Bonferroni = p.adjust(view_df$p_value, method = "bonferroni")

# Visualize
view_df %>%
  arrange(p_value)
```

After Benjamini-Hochberg and Bonferroni correction, all indirect effects are still statistically significant at the 0.05 level.

## Visualization

```{r figures_primary}
est_df = rbind(
  primary_df %>% 
    select(Comparison, Outcome, indirect, CI_LB, CI_UB) %>%
    rename(Est = indirect) %>%
    mutate(Measure = "Indirect"), 
  primary_df %>%
    select(Comparison, Outcome, direct, total) %>%
    pivot_longer(-c(Comparison, Outcome), names_to = "Measure", values_to = "Est") %>%
    mutate(Measure = case_when(Measure == "direct" ~ "Direct", 
                               TRUE ~ "Total"), 
           CI_LB = NA, CI_UB = NA)
  ) %>%
  mutate(Measure = factor(Measure, levels = c("Total", "Direct", "Indirect")))

pd = position_dodge(width = 0.5, orientation = "y")
ref_lines = data.frame(Outcome = out_names, 
                       xint = c(0,1,0,0,0))
est_df %>%
  mutate(Est = case_when(Outcome == "Imposing physical limitations" ~ exp(Est), 
                         TRUE ~ Est), 
         CI_UB = case_when(Outcome == "Imposing physical limitations" ~ exp(CI_UB), 
                         TRUE ~ CI_UB),
         CI_LB = case_when(Outcome == "Imposing physical limitations" ~ exp(CI_LB), 
                         TRUE ~ CI_LB)) %>%
  ggplot(aes(x = Est, y = Comparison, group = Measure, color = Measure)) + 
  geom_point(position = pd) +
  geom_errorbar(aes(xmin = CI_LB, xmax = CI_UB), width = 0.1, 
                position = pd) +
  geom_vline(data = ref_lines, aes(xintercept = xint), linetype = "dashed") + 
  theme_bw() + 
  theme(axis.text.y = element_text(angle = 45, vjust = 0.5, hjust = 1), 
        plot.caption = element_textbox_simple(margin = margin(t = 10)), 
        legend.position = "bottom") + 
  facet_wrap(.~Outcome, scales = "free_x", nrow = 1) + 
  labs(x = "Effect Sizes", y = "Comparison Group", caption = "Note that x-axes are linear for all outcomes other than Imposing physical limitations, for which the effects are expressed as odds ratios.", 
       color = "")
```

# Secondary Analyses: OA Mindset Mediation in the Context of OA Knowledge

We can re-run these same analyses including OA knowledge as a co-mediator to assess mediation of mindset given a particular change in OA knowledge also attributable to the treatment

```{r mediate_secondary, message = FALSE}
model_comps = c("med_OAcomp_No", "med_OAcomp_Edu", 
                "med_KOAS_No", "med_KOAS_Edu", 
                "total_both_No", "total_both_Edu")

build_secondary = function(outcome){
  mod_spec = paste0('
  # Mediator models
  t2_OAcomp ~ a1 * No + a2 * Edu + b * t1_OAcomp
  t2_KOAS ~ c1 * No + c2 * Edu + d * t1_KOAS
  
  # Outcome model
  ', outcome, '~ e1 * No + e2 * Edu + f1 * t2_OAcomp + f2 * t1_OAcomp + g1 * t2_KOAS + g2 * t1_KOAS
  
  # Indirect effects
  med_OAcomp_No := a1 * f1
  med_OAcomp_Edu := a2 * f1
  
  med_KOAS_No := c1 * g1
  med_KOAS_Edu := c2 * g1
  
  # Total effects
  total_both_No := e1 + a1 * f1 + c1 * g1
  total_both_Edu := e2 + a2 * f1 + c2 * g1
  
  # Allow correlation between mediators
  t2_OAcomp ~~ t2_KOAS
  ')
  
  return(mod_spec)
}

secondary_df = data.frame()
for(idx in 1:length(outcomes)){
  # Variable and corresponding SEM model
  varname = outcomes[idx]
  model_spec = build_secondary(varname)
  
  # Create treatment group dummy variables
  fit_df = df %>%
    mutate(
      Edu = case_when(group == "Education" ~ 1, TRUE ~ 0),
      No = case_when(group == "No-intervention" ~ 1, TRUE ~ 0)
    )
  
  # Separate if binary outcome
  if(varname == "Y_limit"){
    fit_df = fit_df %>%
      filter(!is.na(Y_limit)) %>%
      mutate(Y_limit = ordered(Y_limit))
    
    fit = sem(
      model = model_spec,
      data = fit_df,
      se = "robust"
    )
  }
  else{
    fit = sem(
      model = model_spec,
      data = fit_df,
      parallel = "multicore",
      ncpus = 8,
      se = "boot",
      bootstrap = 5000
    )
  }
  
  outcome_df = parameterEstimates(fit, boot.ci.type = "bca.simple") %>%
    filter(label %in% model_comps) %>%
    select(label, est, pvalue, ci.lower, ci.upper) %>%
    separate_wider_delim(label, delim = "_", names = c("Type", "Mediator", "Comparison")) %>%
    mutate(Comparison = case_when(Comparison == "No" ~ "No-intervention", 
                                  TRUE ~ "Education"), 
           Effect = case_when(Mediator == "OAcomp" ~ "OA Mindset", 
                              Mediator == "KOAS" ~ "OA Knowledge", 
                              TRUE ~ "Total"), 
           Outcome = out_names[idx],
           est = -1 * est, CI_LB = -1 * ci.upper, CI_UB = -1 * ci.lower) %>%
    select(Comparison, Outcome, Effect, est, CI_LB, CI_UB, pvalue)
  
  secondary_df = rbind(secondary_df, outcome_df)
}

secondary_df %>%
  arrange(Effect, Comparison, Outcome)
```

Even in the presence of mediation from changes in OA knowledge, there is a statistically significant indirect effect indicating mediation of OA mindset on all of the outcomes except the perceived need for surgery.

## Multiple Testing Adjustment

We can perform a simple multiple testing adjustment of the 10 raw p-values so far estimated for each potential mediator, grouped together as part of examining the potential exercise mindset/OA knowledge mediation of the treatment effect(s) the chosen outcomes. 

```{r mult_adjust_secondary}
# Join outputs of the previous two analyses
view_df_OAcomp = secondary_df %>%
  filter(Effect == "OA Mindset") %>%
  select(Comparison, Outcome, Effect, pvalue)

view_df_OAknow = secondary_df %>%
  filter(Effect == "OA Knowledge") %>%
  select(Comparison, Outcome, Effect, pvalue)

# Adjust the p-values using Benjamini-Hochberg and Bonferroni
view_df_OAcomp$BH = p.adjust(view_df_OAcomp$pvalue, method = "BH")
view_df_OAcomp$Bonferroni = p.adjust(view_df_OAcomp$pvalue, method = "bonferroni")

view_df_OAknow$BH = p.adjust(view_df_OAknow$pvalue, method = "BH")
view_df_OAknow$Bonferroni = p.adjust(view_df_OAknow$pvalue, method = "bonferroni")

# Visualize
rbind(view_df_OAknow, view_df_OAcomp) %>%
  arrange(Effect, Comparison, Outcome)
```

After Benjamini-Hochberg and Bonferroni correction, indirect effects for the mindset mediation of treatment effect on imposing physical limitations lose statistical significant at the 0.05 level vs both of the other arms. However, all other indirect effects that were previously significant are still statistically significant at the 0.05 level.

## Visualization

```{r figures_secondary}
est_df = secondary_df %>%
  select(-c(pvalue)) %>%
  mutate(CI_LB = case_when(Effect == "OA Mindset" ~ CI_LB, 
                              TRUE ~ NA), 
         CI_UB = case_when(Effect == "OA Mindset" ~ CI_UB, 
                              TRUE ~ NA),
         Effect = factor(Effect, levels = c("Total", "OA Knowledge", "OA Mindset"))) %>%
  rename(Est = est, Measure = Effect)

pd = position_dodge(width = 0.5, orientation = "y")
ref_lines = data.frame(Outcome = out_names, 
                       xint = c(0,1,0,0,0))
est_df %>%
  mutate(Est = case_when(Outcome == "Imposing physical limitations" ~ exp(Est), 
                         TRUE ~ Est), 
         CI_UB = case_when(Outcome == "Imposing physical limitations" ~ exp(CI_UB), 
                         TRUE ~ CI_UB),
         CI_LB = case_when(Outcome == "Imposing physical limitations" ~ exp(CI_LB), 
                         TRUE ~ CI_LB)) %>%
  ggplot(aes(x = Est, y = Comparison, group = Measure, color = Measure)) + 
  geom_point(position = pd) +
  geom_errorbar(aes(xmin = CI_LB, xmax = CI_UB), width = 0.1, 
                position = pd) +
  geom_vline(data = ref_lines, aes(xintercept = xint), linetype = "dashed") + 
  theme_bw() + 
  theme(axis.text.y = element_text(angle = 45, vjust = 0.5, hjust = 1), 
        plot.caption = element_textbox_simple(margin = margin(t = 10)), 
        legend.position = "bottom") + 
  facet_wrap(.~Outcome, scales = "free_x", nrow = 1) + 
  labs(x = "Effect Sizes", y = "Comparison Group", caption = "Note that x-axes are linear for all outcomes other than Imposing physical limitations, for which the effects are expressed as odds ratios.", 
       color = "")
```
