---
title: "Treatment Pathways"
author: "Joe Sartini"
date: "2026-09-01"
output:
  html_document: default
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set-up

First, all necessary libraries are imported, the data is read in, and all necessary variables are produced.

```{r set_up, message=FALSE}
# First, the libraries 
# If you want to replicate my environment exactly, use renv::restore()
library(tidyverse)
library(mediation)
library(lavaan)
library(knitr)
library(ggtext)
library(corrplot)
select = dplyr::select

# Next, reading the data
GLAB = c("Mindset", "Education", "No-intervention")
df = read.csv("../deidentified_data.csv") %>%
  mutate(group = recode_values(t2_Group.ID, from = c(1, 2, 3),
                               to = GLAB), # Group names
         group = factor(group, levels = GLAB[c(3,2,1)]),
         female = case_when(t1_sex == 1 ~ 1,
                            t1_sex == 0 ~ 0,
                            TRUE ~ NA), # Recode sex
         t3_OAcomp = t3_IMI_man + t3_IMI_opp - t3_IMI_cat,
         t2_OAcomp = t2_IMI_man + t2_IMI_opp - t2_IMI_cat,
         t1_OAcomp = t1_IMI_man + t1_IMI_opp - t1_IMI_cat,
         bilateral = case_when(t1_uni.bi == 5 ~ 1,
                               t1_uni.bi == 4 ~ 0,
                               TRUE ~ NA),
         minority = case_when(!(t1_race %in% c("6", "")) ~ 1,
                              TRUE ~ 0),
         employed = case_when(grepl("[1-2]", t1_employment) ~ 1,
                              TRUE ~ 0),
         depression = case_when(grepl("10", t1_med.cond) ~ 1, 
                                TRUE ~ 0),
         n_comorbid = lengths(strsplit(t1_med.cond, ",")))
```

# OA Mindset Mediation of Treatment Effects on 5 Outcomes

## Mediation

We perform separate causal mediation analyses to assess mindset change (time point 2 - time point 1) as a potential mediator of the treatment effect on 5 outcomes which were impacted by the treatment. These outcomes include difference in perceived need for surgery (time point 3 - time point 1), removal of self-imposed physical limitation (time point 1 to time point 3), fear of movement change, change in pain self-efficacy (time point 3 - time point 1), and change in other symptom self-efficacy (time point 3 - time point 1). 

We assess mediation of the movement from both the no intervention and education arms to the mindset treatment arm. Generalized linear models are used to model the binary removal of self-imposed physical limitation outcome. Uncertainty is assessed using bootstrap inference.

```{r mediate_single, message = FALSE}
outcomes = c("surgery", "management_11", "TSK_avg", "ASE_pain_avg", "ASE_function_avg")

out_names = c("Perceived need for surgery",
              "Imposing physical limitations", 
              "Fear of movement",
              "Pain self-efficacy", 
              "Other symptom self-efficacy")

build_ind = function(outcome){
  if(outcome == "management_11"){
    mod_spec = paste0(
      '
  # Mediator Model at time 3
  t3_OAcomp ~ t2_OAcomp

  # Mediator Model at time 2
  t2_OAcomp ~ a1 * No + a2 * Edu + t1_OAcomp

  # Outcome models
  t3_',
  outcome,
  '~ c1 * No + c2 * Edu + d * t2_OAcomp

  # Indirect effects
  med_OAcomp_No := a1 * d
  med_OAcomp_Edu := a2 * d

  # Total effects
  total_both_No := c1 + (a1 * d)
  total_both_Edu := c2 + (a2 * d)
  '
    )
  }
  else{
    mod_spec = paste0(
      '
  # Mediator Model at time 3
  t3_OAcomp ~ t2_OAcomp

  # Mediator Model at time 2
  t2_OAcomp ~ a1 * No + a2 * Edu + t1_OAcomp

  # Outcome models
  t3_',
  outcome,
  '~ c1 * No + c2 * Edu + d * t2_OAcomp + t1_',
  outcome,
  '

  # Indirect effects
  med_OAcomp_No := a1 * d
  med_OAcomp_Edu := a2 * d

  # Total effects
  total_both_No := c1 + (a1 * d)
  total_both_Edu := c2 + (a2 * d)
  '
    )
  }
  
  return(mod_spec)
}

ind_df = data.frame()
for(idx in 1:length(outcomes)){
  # Variable and corresponding SEM model
  varname = outcomes[idx]
  model_spec = build_ind(varname)
  
  # Create treatment group dummy variables
  fit_df = df %>%
    mutate(
      Edu = case_when(group == "Education" ~ 1, TRUE ~ 0),
      No = case_when(group == "No-intervention" ~ 1, TRUE ~ 0)
    )
  
  # Separate if binary outcome
  if(varname == "management_11"){
    fit_df = fit_df %>% 
      filter(t1_management_11 == 1) %>%
      mutate(t3_management_11 = ordered(t3_management_11, levels = c(0, 1)))
    
    fit = sem(
      model = model_spec,
      data = fit_df,
      se = "robust"
    )
  }
  else{
    fit = sem(
      model = model_spec,
      data = fit_df,
      parallel = "multicore",
      ncpus = 8,
      se = "boot",
      bootstrap = 5000
    )
  }
  
  outcome_df = parameterEstimates(fit, boot.ci.type = "bca.simple") %>%
    filter(label %in% c("med_OAcomp_No", "med_OAcomp_Edu",
                        "total_both_No", "total_both_Edu")) %>%
    select(label, est, pvalue, ci.lower, ci.upper) %>%
    separate_wider_delim(label, delim = "_", names = c("Type", "Mediator", "Comparison")) %>%
    mutate(Comparison = case_when(Comparison == "No" ~ "No-intervention", 
                                  TRUE ~ "Education"), 
           Effect = case_when(Mediator == "OAcomp" ~ "OA Mindset", 
                              TRUE ~ "Total"), 
           Outcome = out_names[idx],
           est = -1 * est, CI_LB = -1 * ci.upper, CI_UB = -1 * ci.lower) %>%
    select(Comparison, Outcome, Effect, est, CI_LB, CI_UB, pvalue)
  
  ind_df = rbind(ind_df, outcome_df)
}

ind_df %>%
  arrange(Effect, Comparison, Outcome)
```

We see that mindset mediates the treatment effects for both contrasts and all outcomes quite strongly. Note that the "percent_mediated" column just reports the relative size of the mediation effect relative to the total effect, so it can fall outside of the 0\%-100\% range. 

## Multiple Testing Adjustment

We can perform a simple multiple testing adjustment of the 10 raw p-values so far estimated, grouped together as part of examining the potential exercise mindset mediation of the treatment effect(s) the chosen outcomes. 

```{r mult_adjust_single}
# Restrict just to the mindset indirect effects
view_df = ind_df %>%
  filter(Effect == "OA Mindset") %>%
  select(Comparison, Outcome, pvalue)

# Adjust the p-values using Benjamini-Hochberg and Bonferroni
view_df$BH = p.adjust(view_df$pvalue, method = "BH")
view_df$Bonferroni = p.adjust(view_df$pvalue, method = "bonferroni")

# Visualize
view_df %>%
  arrange(pvalue)
```

After Benjamini-Hochberg and Bonferroni correction, all indirect effects are still statistically significant at the 0.05 level.

## Visualization

```{r figures_single}
est_df = ind_df %>%
  select(-c(pvalue)) %>%
  mutate(Effect = factor(Effect, levels = c("Total", "OA Mindset"))) %>%
  rename(Est = est, Measure = Effect) %>%
  mutate(CI_UB = case_when(Measure == "Total" ~ NA, 
                           TRUE ~ CI_UB), 
         CI_LB = case_when(Measure == "Total" ~ NA, 
                           TRUE ~ CI_LB))

pd = position_dodge(width = 0.5, orientation = "y")
ref_lines = data.frame(Outcome = out_names, 
                       xint = c(0,1,0,0,0))
est_df %>%
  mutate(Est = case_when(Outcome == "Imposing physical limitations" ~ exp(Est), 
                         TRUE ~ Est), 
         CI_UB = case_when(Outcome == "Imposing physical limitations" ~ exp(CI_UB), 
                         TRUE ~ CI_UB),
         CI_LB = case_when(Outcome == "Imposing physical limitations" ~ exp(CI_LB), 
                         TRUE ~ CI_LB)) %>%
  ggplot(aes(x = Est, y = Comparison, color = Measure, group = Measure)) + 
  geom_point(position = pd) +
  geom_errorbar(aes(xmin = CI_LB, xmax = CI_UB), width = 0.1, 
                position = pd) +
  geom_vline(data = ref_lines, aes(xintercept = xint), linetype = "dashed") + 
  theme_bw() + 
  theme(axis.text.y = element_text(angle = 45, vjust = 0.5, hjust = 1), 
        plot.caption = element_textbox_simple(margin = margin(t = 10)), 
        legend.position = "bottom") + 
  facet_wrap(.~Outcome, scales = "free_x", nrow = 1) + 
  labs(x = "Effect Sizes", y = "Comparison Group", caption = "Note that x-axes are linear for all outcomes other than Imposing physical limitations, for which the effects are expressed as odds ratios.", 
       color = "")
```

# OA Mindset and OA Knowledge Joint Mediation

We can re-run these same analyses including OA knowledge as a co-mediator to assess mediation of mindset given a particular change in OA knowledge also attributable to the treatment

```{r mediate_joint, message = FALSE}
outcomes = c("surgery", "management_11", "TSK_avg", "ASE_pain_avg", "ASE_function_avg")

out_names = c("Perceived need for surgery",
              "Imposing physical limitations", 
              "Fear of movement",
              "Pain self-efficacy", 
              "Other symptom self-efficacy")

build_joint = function(outcome){
  if(outcome == "management_11"){
    mod_spec = paste0(
      '
  # Autoregressive models
  t3_OAcomp ~ t2_OAcomp
  t3_KOAS ~ t2_KOAS

  # Mediation Models
  t2_OAcomp ~ a1 * No + a2 * Edu + t1_OAcomp
  t2_KOAS ~ b1 * No + b2 * Edu + t1_KOAS

  # Outcome models
  t3_',
  outcome,
  '~ e1 * No + e2 * Edu + f * t2_OAcomp + g * t2_KOAS

  # Indirect effects
  med_OAcomp_No := a1 * f
  med_OAcomp_Edu := a2 * f

  med_KOAS_No := b1 * g
  med_KOAS_Edu := b2 * g

  # Total effects
  total_both_No := e1 + a1 * f + b1 * g
  total_both_Edu := e2 + a2 * f + b2 * g

  # Allow residual correlation of mediators
  t2_OAcomp ~~ t2_KOAS
  t3_OAcomp ~~ t3_KOAS
  '
    )
  }
  else{
    mod_spec = paste0(
      '
  # Autoregressive models
  t3_OAcomp ~ t2_OAcomp
  t3_KOAS ~ t2_KOAS

  # Mediation Models
  t2_OAcomp ~ a1 * No + a2 * Edu + t1_OAcomp
  t2_KOAS ~ b1 * No + b2 * Edu + t1_KOAS

  # Outcome models
  t3_',
  outcome,
  '~ e1 * No + e2 * Edu + f * t2_OAcomp + g * t2_KOAS + t1_',
  outcome,
  '

  # Indirect effects
  med_OAcomp_No := a1 * f
  med_OAcomp_Edu := a2 * f

  med_KOAS_No := b1 * g
  med_KOAS_Edu := b2 * g

  # Total effects
  total_both_No := e1 + a1 * f + b1 * g
  total_both_Edu := e2 + a2 * f + b2 * g

  # Allow residual correlation of mediators
  t2_OAcomp ~~ t2_KOAS
  t3_OAcomp ~~ t3_KOAS
  '
    )
  }
  
  
  return(mod_spec)
}

joint_df = data.frame()
for(idx in 1:length(outcomes)){
  # Variable and corresponding SEM model
  varname = outcomes[idx]
  model_spec = build_joint(varname)
  
  # Create treatment group dummy variables
  fit_df = df %>%
    mutate(
      Edu = case_when(group == "Education" ~ 1, TRUE ~ 0),
      No = case_when(group == "No-intervention" ~ 1, TRUE ~ 0)
    )
  
  # Separate if binary outcome
  if(varname == "management_11"){
    fit_df = fit_df %>% 
      filter(t1_management_11 == 1) %>%
      mutate(t3_management_11 = ordered(t3_management_11, levels = c(0, 1)))
    
    fit = sem(
      model = model_spec,
      data = fit_df,
      se = "robust"
    )
  }
  else{
    fit = sem(
      model = model_spec,
      data = fit_df,
      parallel = "multicore",
      ncpus = 8,
      se = "boot",
      bootstrap = 5000
    )
  }
  
  outcome_df = parameterEstimates(fit, boot.ci.type = "bca.simple") %>%
    filter(label %in% c("med_OAcomp_No", "med_OAcomp_Edu", 
                        "med_KOAS_No", "med_KOAS_Edu", 
                        "total_both_No", "total_both_Edu")) %>%
    select(label, est, pvalue, ci.lower, ci.upper) %>%
    separate_wider_delim(label, delim = "_", names = c("Type", "Mediator", "Comparison")) %>%
    mutate(Comparison = case_when(Comparison == "No" ~ "No-intervention", 
                                  TRUE ~ "Education"), 
           Effect = case_when(Mediator == "OAcomp" ~ "OA Mindset", 
                              Mediator == "KOAS" ~ "OA Knowledge", 
                              TRUE ~ "Total"), 
           Outcome = out_names[idx],
           est = -1 * est, CI_LB = -1 * ci.upper, CI_UB = -1 * ci.lower) %>%
    select(Comparison, Outcome, Effect, est, CI_LB, CI_UB, pvalue)
  
  joint_df = rbind(joint_df, outcome_df)
}

joint_df %>%
  arrange(Effect, Comparison, Outcome)
```

Even in the presence of mediation from changes in OA knowledge, there is a statistically significant indirect effect indicating mediation of OA mindset on all of the outcomes except the perceived need for surgery.

## Multiple Testing Adjustment

We can perform a simple multiple testing adjustment of the 10 raw p-values so far estimated for each potential mediator, grouped together as part of examining the potential exercise mindset/OA knowledge mediation of the treatment effect(s) the chosen outcomes. 

```{r mult_adjust_joint}
# Join outputs of the previous two analyses
view_df_OAcomp = joint_df %>%
  filter(Effect == "OA Mindset") %>%
  select(Comparison, Outcome, Effect, pvalue)

view_df_OAknow = joint_df %>%
  filter(Effect == "OA Knowledge") %>%
  select(Comparison, Outcome, Effect, pvalue)

# Adjust the p-values using Benjamini-Hochberg and Bonferroni
view_df_OAcomp$BH = p.adjust(view_df_OAcomp$pvalue, method = "BH")
view_df_OAcomp$Bonferroni = p.adjust(view_df_OAcomp$pvalue, method = "bonferroni")

view_df_OAknow$BH = p.adjust(view_df_OAknow$pvalue, method = "BH")
view_df_OAknow$Bonferroni = p.adjust(view_df_OAknow$pvalue, method = "bonferroni")

# Visualize
rbind(view_df_OAknow, view_df_OAcomp) %>%
  arrange(Effect, Comparison, Outcome)
```

After Benjamini-Hochberg and Bonferroni correction, indirect effects for the mindset mediation of treatment effect on imposing physical limitations lose statistical significant at the 0.05 level vs both of the other arms. However, all other indirect effects that were previously significant are still statistically significant at the 0.05 level.

## Visualization

```{r figures_joint}
est_df = joint_df %>%
  select(-c(pvalue)) %>%
  mutate(Effect = factor(Effect, levels = c("Total", "OA Knowledge", "OA Mindset"))) %>%
  rename(Est = est, Measure = Effect) %>%
  filter(Measure != "Total")

pd = position_dodge(width = 0.5, orientation = "y")
ref_lines = data.frame(Outcome = out_names, 
                       xint = c(0,1,0,0,0))
est_df %>%
  mutate(Est = case_when(Outcome == "Imposing physical limitations" ~ exp(Est), 
                         TRUE ~ Est), 
         CI_UB = case_when(Outcome == "Imposing physical limitations" ~ exp(CI_UB), 
                         TRUE ~ CI_UB),
         CI_LB = case_when(Outcome == "Imposing physical limitations" ~ exp(CI_LB), 
                         TRUE ~ CI_LB)) %>%
  ggplot(aes(x = Est, y = Comparison, group = Measure, color = Measure)) + 
  geom_point(position = pd) +
  geom_errorbar(aes(xmin = CI_LB, xmax = CI_UB), width = 0.1, 
                position = pd) +
  geom_vline(data = ref_lines, aes(xintercept = xint), linetype = "dashed") + 
  theme_bw() + 
  theme(axis.text.y = element_text(angle = 45, vjust = 0.5, hjust = 1), 
        plot.caption = element_textbox_simple(margin = margin(t = 10)), 
        legend.position = "bottom") + 
  facet_wrap(.~Outcome, scales = "free_x", nrow = 1) + 
  labs(x = "Effect Sizes", y = "Comparison Group", caption = "Note that x-axes are linear for all outcomes other than Imposing physical limitations, for which the effects are expressed as odds ratios.", 
       color = "")
```

# Exploratory Latent Factor Models

To understand what might be leading to the impact of both mindset and the education arms on physical activity, we fit mediation models which include latent factors aiming to incorporate mindset variables, fear of movement, and self-efficacy variates collected at the various time points. We begin by examining how many latent factors will be appropriate for this data. We proceed under the assumption that the factor weights will be consistent over time.

```{r latent_numfact, warning=FALSE}
factor_vars = c(paste0("t", 1:3, "_OAcomp"), 
                paste0("t", 1:3, "_KOAS"), 
                paste0("t", 1:3, "_PEM_avg"), 
                paste0("t", 1:3, "_TSK_avg"), 
                paste0("t", 1:3, "_ASE_pain_avg"), 
                paste0("t", 1:3, "_ASE_function_avg"), 
                paste0("t", c(1,3), "_PASE_total"))

# Rescale all variables for numerical stability 0-10 scale
{
  max_OAcomp = max(df %>% select(all_of(factor_vars[c(1,2,3)])) %>% as.matrix() %>% c()) / 10
  max_KOAS = max(df %>% select(all_of(factor_vars[c(4,5,6)])) %>% as.matrix() %>% c()) / 10
  max_PEM = max(df %>% select(all_of(factor_vars[c(7,8,9)])) %>% as.matrix() %>% c()) / 10
  max_TSK = max(df %>% select(all_of(factor_vars[c(10,11,12)])) %>% as.matrix() %>% c()) / 10
  max_pain = max(df %>% select(all_of(factor_vars[c(13,14,15)])) %>% as.matrix() %>% c()) / 10
  max_func = max(df %>% select(all_of(factor_vars[c(16,17,18)])) %>% as.matrix() %>% c()) / 10
  max_PASE = max(df %>% select(all_of(factor_vars[c(19,20)])) %>% as.matrix() %>% c()) / 10
}

# Data for latent modeling
latent_df = df %>%
  mutate(
    Edu = case_when(group == "Education" ~ 1, TRUE ~ 0),
    No = case_when(group == "No-intervention" ~ 1, TRUE ~ 0),
    across(all_of(factor_vars[c(1,2,3)]), ~ .x/max_OAcomp), 
    across(all_of(factor_vars[c(4,5,6)]), ~ .x/max_KOAS),
    across(all_of(factor_vars[c(7,8,9)]), ~ .x/max_PEM),
    across(all_of(factor_vars[c(10,11,12)]), ~ .x/max_TSK),
    across(all_of(factor_vars[c(13,14,15)]), ~ .x/max_pain),
    across(all_of(factor_vars[c(16,17,18)]), ~ .x/max_func),
    across(all_of(factor_vars[c(19,20)]), ~ .x/max_PASE))  %>%
  select(all_of(c(factor_vars, "Edu", "No", "group")))

# ELFA models
factor_1 = '
  # Latent factor - time 3
  f1_t3 =~ l11 * t3_OAcomp + l12 * t3_KOAS + l13 * t3_PEM_avg +
           l14 * t3_TSK_avg + l15 * t3_ASE_pain_avg + l16 * t3_ASE_function_avg

  # Latent factor - time 2
  f1_t2 =~ l11 * t2_OAcomp + l12 * t2_KOAS + l13 * t2_PEM_avg +
           l14 * t2_TSK_avg + l15 * t2_ASE_pain_avg + l16 * t2_ASE_function_avg

  # Latent factor - time 1
  f1_t1 =~ l11 * t1_OAcomp + l12 * t1_KOAS + l13 * t1_PEM_avg +
           l14 * t1_TSK_avg + l15 * t1_ASE_pain_avg + l16 * t1_ASE_function_avg
  
  # Correlated residuals (factors do not capture all autocorrelated variability)
  t3_OAcomp ~~ t2_OAcomp + t1_OAcomp; t2_OAcomp ~~ t1_OAcomp
  t3_KOAS ~~ t2_KOAS + t1_KOAS; t2_KOAS ~~ t1_KOAS
  t3_PEM_avg ~~ t2_PEM_avg + t1_PEM_avg; t2_PEM_avg ~~ t1_PEM_avg
  t3_TSK_avg ~~ t2_TSK_avg + t1_TSK_avg; t2_TSK_avg ~~ t1_TSK_avg
  t3_ASE_pain_avg ~~ t2_ASE_pain_avg + t1_ASE_pain_avg; t2_ASE_pain_avg ~~ t1_ASE_pain_avg
  t3_ASE_function_avg ~~ t2_ASE_function_avg + t1_ASE_function_avg; t2_ASE_function_avg ~~ t1_ASE_function_avg
  
  # Latent factor dependence over time, includes treatment effect
  f1_t3 ~ f1_t2; f1_t2 ~ f1_t1 + No + Edu
'

factor_2 = '
  # Latent factor - time 3
  f1_t3 =~ l11 * t3_OAcomp + l12 * t3_KOAS + l13 * t3_PEM_avg +
           l14 * t3_TSK_avg + l15 * t3_ASE_pain_avg + l16 * t3_ASE_function_avg
           
  f2_t3 =~ l21 * t3_OAcomp + l22 * t3_KOAS + l23 * t3_PEM_avg +
           l24 * t3_TSK_avg + l25 * t3_ASE_pain_avg + l26 * t3_ASE_function_avg

  # Latent factor - time 2
  f1_t2 =~ l11 * t2_OAcomp + l12 * t2_KOAS + l13 * t2_PEM_avg +
           l14 * t2_TSK_avg + l15 * t2_ASE_pain_avg + l16 * t2_ASE_function_avg
           
  f2_t2 =~ l21 * t2_OAcomp + l22 * t2_KOAS + l23 * t2_PEM_avg +
           l24 * t2_TSK_avg + l25 * t2_ASE_pain_avg + l26 * t2_ASE_function_avg

  # Latent factor - time 1
  f1_t1 =~ l11 * t1_OAcomp + l12 * t1_KOAS + l13 * t1_PEM_avg +
           l14 * t1_TSK_avg + l15 * t1_ASE_pain_avg + l16 * t1_ASE_function_avg
           
  f2_t1 =~ l21 * t1_OAcomp + l22 * t1_KOAS + l23 * t1_PEM_avg +
           l24 * t1_TSK_avg + l25 * t1_ASE_pain_avg + l26 * t1_ASE_function_avg
  
  # Correlated residuals (factors do not capture all autocorrelated variability)
  t3_OAcomp ~~ t2_OAcomp + t1_OAcomp; t2_OAcomp ~~ t1_OAcomp
  t3_KOAS ~~ t2_KOAS + t1_KOAS; t2_KOAS ~~ t1_KOAS
  t3_PEM_avg ~~ t2_PEM_avg + t1_PEM_avg; t2_PEM_avg ~~ t1_PEM_avg
  t3_TSK_avg ~~ t2_TSK_avg + t1_TSK_avg; t2_TSK_avg ~~ t1_TSK_avg
  t3_ASE_pain_avg ~~ t2_ASE_pain_avg + t1_ASE_pain_avg; t2_ASE_pain_avg ~~ t1_ASE_pain_avg
  t3_ASE_function_avg ~~ t2_ASE_function_avg + t1_ASE_function_avg; t2_ASE_function_avg ~~ t1_ASE_function_avg
  
  # Latent factor dependence over time, includes treatment effect
  f1_t3 ~ f1_t2; f1_t2 ~ f1_t1 + No + Edu
  f2_t3 ~ f2_t2; f2_t2 ~ f2_t1 + No + Edu
'
# Fit both models
fit_1 = cfa(factor_1, data = latent_df)
fit_2 = cfa(factor_2, data = latent_df)

# Model with two factors is statistically superior
lavTestLRT(fit_1, fit_2) 

# Second factor has no well-determined loadings
parameterEstimates(fit_2) %>%
  subset(op == "=~")

# Single factor model provides similar goodness of fit
summary(fit_1, fit.measures = T)
summary(fit_2, fit.measures = T)
```

The above analyses indicate a slightly better fit for the model with two factors, as well as an acceptable representation of the data for this model according to standard goodness of fit metrics. As a result, we proceed with a very simple model using two latent factors.

Examination of the latent factor loadings above indicates that the second factor is not well-defined (no factor loadings are statistically significant). This leads to issues with bootstrap resampling stability, and so we switch here to use of a robust maximum likelihood procedure.

```{r latent_mediation}
# Mediation and first factor fit
latent_mod = '
   # Latent factor - time 3
  f1_t3 =~ l11 * t3_OAcomp + l12 * t3_KOAS + l13 * t3_PEM_avg +
           l14 * t3_TSK_avg + l15 * t3_ASE_pain_avg + l16 * t3_ASE_function_avg
           
  f2_t3 =~ l21 * t3_OAcomp + l22 * t3_KOAS + l23 * t3_PEM_avg +
           l24 * t3_TSK_avg + l25 * t3_ASE_pain_avg + l26 * t3_ASE_function_avg

  # Latent factor - time 2
  f1_t2 =~ l11 * t2_OAcomp + l12 * t2_KOAS + l13 * t2_PEM_avg +
           l14 * t2_TSK_avg + l15 * t2_ASE_pain_avg + l16 * t2_ASE_function_avg
           
  f2_t2 =~ l21 * t2_OAcomp + l22 * t2_KOAS + l23 * t2_PEM_avg +
           l24 * t2_TSK_avg + l25 * t2_ASE_pain_avg + l26 * t2_ASE_function_avg

  # Latent factor - time 1
  f1_t1 =~ l11 * t1_OAcomp + l12 * t1_KOAS + l13 * t1_PEM_avg +
           l14 * t1_TSK_avg + l15 * t1_ASE_pain_avg + l16 * t1_ASE_function_avg
           
  f2_t1 =~ l21 * t1_OAcomp + l22 * t1_KOAS + l23 * t1_PEM_avg +
           l24 * t1_TSK_avg + l25 * t1_ASE_pain_avg + l26 * t1_ASE_function_avg
  
  # Correlated residuals (factors do not capture all autocorrelated variability)
  t3_OAcomp ~~ t2_OAcomp + t1_OAcomp; t2_OAcomp ~~ t1_OAcomp
  t3_KOAS ~~ t2_KOAS + t1_KOAS; t2_KOAS ~~ t1_KOAS
  t3_PEM_avg ~~ t2_PEM_avg + t1_PEM_avg; t2_PEM_avg ~~ t1_PEM_avg
  t3_TSK_avg ~~ t2_TSK_avg + t1_TSK_avg; t2_TSK_avg ~~ t1_TSK_avg
  t3_ASE_pain_avg ~~ t2_ASE_pain_avg + t1_ASE_pain_avg; t2_ASE_pain_avg ~~ t1_ASE_pain_avg
  t3_ASE_function_avg ~~ t2_ASE_function_avg + t1_ASE_function_avg; t2_ASE_function_avg ~~ t1_ASE_function_avg
  
  # Latent factor dependence over time
  f1_t3 ~ f1_t2
  f2_t3 ~ f2_t2
  
  # Mediation models, includes treatment effect
  f1_t2 ~ a1 * No + a2 * Edu + f1_t1
  f2_t2 ~ b1 * No + b2 * Edu + f2_t1
  
  # Outcome model
  t3_PASE_total ~ c1 * No + c2 * Edu + d * f1_t2 + t1_PASE_total + e * f2_t2 
  
  # Indirect effects
  med_F1_No := a1 * d
  med_F1_Edu := a2 * d
  
  med_F2_No := b1 * e
  med_F2_Edu := b2 * e

  # Total effects
  total_B_No := c1 + a1 * d + b1 * e
  total_B_Edu := c2 + a2 * d + b2 * e 
'

latent_fit = sem(
  model = latent_mod,
  data = latent_df,
  estimator = "MLR"
)

# Latent factor mediation
parameterEstimates(latent_fit) %>% # , boot.ci.type = "bca.simple") %>%
  filter(label %in% c("med_F1_No", "med_F1_Edu", 
                      "med_F2_No", "med_F2_Edu",
                      "total_B_No", "total_B_Edu")) %>%
  select(label, est, pvalue, ci.lower, ci.upper) %>%
  separate_wider_delim(label,
                       delim = "_",
                       names = c("Type", "Factor", "Comparison")) %>%
  mutate(Effect = case_when(Type == "med" & Factor == "F1" ~ "F1 Indirect", 
                            Type == "med" & Factor == "F2" ~ "F2 Indirect",
                            TRUE ~ "Total Effect"),
         Comparison = case_when(Comparison == "No" ~ "No-intervention", 
                                TRUE ~ "Education"), 
         est = -1 * est, CI_LB = -1 * ci.upper, CI_UB = -1 * ci.lower) %>%
  select(Comparison, Effect, est, CI_LB, CI_UB, pvalue)
```

We find that the indirect effect of the second latent factor is statistically significant for both comparisons. We can now examine the latent factor loadings to understand this factor's components, as well as how the latent factor distributions develop over the course of the treatment period by arm.

```{r latent_loadings}
{
  weights = parameterEstimates(latent_fit, boot.ci.type = "norm") %>%
    filter(label %in% c("l21", "l22", "l23", "l24", "l25", "l26")) %>%
    select(label, est) %>%
    group_by(label) %>%
    summarise(est = first(est)) %>%
    arrange(label) %>%
    pull(est)
  
  names(weights) = c("OA Mindset", "OA Knowledge", "Exercise Mindset", 
                     "Fear of Movement", "Pain Self-Efficacy", "Other Self-Efficacy")
  print(weights)
}
```

We see that the latent factor measures OA mindset and self-efficacy in a similar magnitude and positive direction, while exercise mindset and OA knowledge are in the same direction but lower in magnitude. Fear of movement contributes a similar magnitude to these secondary factors, but in a negative direction.

```{r latent_vis}
latent_plot_df = latent_df %>%
  mutate(latent_1 = weights[1] * t1_OAcomp + 
           weights[2] * t1_KOAS + 
           weights[3] * t1_PEM_avg + 
           weights[4] * t1_TSK_avg + 
           weights[5] * t1_ASE_pain_avg + 
           weights[6] * t1_ASE_function_avg, 
         latent_2 = weights[1] * t2_OAcomp + 
           weights[2] * t2_KOAS + 
           weights[3] * t2_PEM_avg + 
           weights[4] * t2_TSK_avg + 
           weights[5] * t2_ASE_pain_avg + 
           weights[6] * t2_ASE_function_avg, 
         latent_3 = weights[1] * t3_OAcomp + 
           weights[2] * t3_KOAS + 
           weights[3] * t3_PEM_avg + 
           weights[4] * t3_TSK_avg + 
           weights[5] * t3_ASE_pain_avg + 
           weights[6] * t3_ASE_function_avg) %>%
  select(group, latent_1, latent_2, latent_3) %>%
  mutate(ID = row_number()) %>%
  pivot_longer(-c(ID, group), names_to = "Time", values_to = "Latent") %>%
  mutate(Time = case_when(Time == "latent_1" ~ "Time Point 1", 
                          Time == "latent_2" ~ "Time Point 2",
                          TRUE ~"Time Point 3"),
         group = factor(group, levels = c("Mindset", "Education", "No-intervention")))

# Densities at each time point
latent_plot_df %>%
  ggplot(aes(x = Latent, group = group, fill = group)) + 
  geom_density(alpha = 0.5) + 
  theme_classic() + 
  facet_grid(.~Time) + 
  labs(fill = "", x = "Latent Factor", y = "Density")

# Trajectories by group over time
latent_plot_df %>%
  ggplot(aes(x = Time, y = Latent)) + 
  geom_line(aes(group = ID, color = group), alpha = 0.2) +
  geom_smooth(aes(group = group, color = group), method = "lm", 
              se = F) + 
  theme_classic() + 
  labs(x = "Time Point", y = "Latent Factor", color = "", fill = "")
```
